---
title: "Community"
author: "Sarah Heidmann"
date: "9/14/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, include=FALSE}
# Load libraries
library(tidyverse)
library(knitr)
library(vegan)
```
```{r dataImport, include=FALSE}
# Import data
dl <- read_csv("data/DeepLionFish_qaqc_20200413.csv")
# Set colors
mpaColors <- c("#E69F00", "#56B4E9")
inoutColors <- c("#00BE6C","#F8766D")
allcolors <- c("MCD"="#E69F00","VICRNM"="#56B4E9",
               "IN"="#00BE6C","OUT"="#F8766D")
grouplabels <- c("LB.non.pred.non.comp"="Large-bodied non-predator",
                 "LB.predator"="Large-bodied predator",
                 "Non.Reef.Asso"="Non-reef associated",
                 "SB.NonPrey"="Small-bodied non-prey",
                 "SB.Prey"="Small-bodied prey",
                 "Lion"="Lionfish")
```

## Summary

This document examines fish community composition inside and outside the boundaries of the Hind Bank Marine Conservation District (MCD) and Virgin Islands Coral Reef National Monument (VICRNM).

## Density and biomass of groups of interest by site

First we need to calculate the density (#/100m^2^) and biomass (g/100m^2^) of each group at each site.

The code below does the following:

1. Create columns for lionfish
2. Remove unnecessary columns
3. Group by site, MPA, and in/out location
4. Calculate the sum of each group's density and biomass

```{r groupSiteSum}
groupsitesum <- dl %>%
     mutate(Lion=ifelse(ScientificName=="Pterois volitans", TotalNumber, 0),
            Lion.biomass=ifelse(ScientificName=="Pterois volitans", TotalBiomass, 0)) %>%
     select(SiteID,MPA,InOut,
            all_of(names(grouplabels)), 
            all_of(paste0(names(grouplabels),".biomass"))) %>%
     group_by(SiteID, MPA, InOut) %>%
     summarise(across(everything(), sum), .groups="drop")
```

This table is the first 6 rows of the result:

```{r groupSiteSumTab, echo=FALSE}
kable(head(groupsitesum), digits=0, align = "c",
      col.names = c("Site", "MPA","Location",
                    paste0(grouplabels, " Density"), paste0(grouplabels," Biomass")),
      caption = "Total density (#/100m^2^) and biomass (g/100m^2^) of groups of interest across sites (first 6 rows).")
```


### NMDS for density of groups

The code below creates a matrix of densities of groups of interest at each site, using that to create an NMDS and the associated stressplot.

```{r groupDens}
density_matrix <- as.matrix(groupsitesum[names(grouplabels)])
dimnames(density_matrix) <- list(as.character(groupsitesum$SiteID), names(grouplabels))
#dens_dist <- vegdist(density_matrix, method="jaccard")
dens_NMDS <- metaMDS(density_matrix, distance="jaccard", k=2)
stressplot(dens_NMDS)
```

The fit is good with fairly low stress.

The following code converts the NMDS into a data frame and plots it.

```{r groupDensPlot, fig.width = 7.5, fig.height = 7, fig.cap="NMDS plot of density of various functional groups of interest by MPA. Groups were adapted from Green et al (2012). Ellipses represent 95% groupings by MPA (inside and outside combined for each MPA)."}
dens_sites <- as.data.frame(scores(dens_NMDS))
dens_sites$site <- rownames(dens_sites)
dens_sites$mpa <- groupsitesum$MPA
dens_sites$inout <- groupsitesum$InOut

dens_species <- as.data.frame(scores(dens_NMDS, display="species"))
dens_species$species <- rownames(dens_species)
dens_species$specieslabels <- grouplabels

ggplot() + 
   geom_label(data=dens_species,aes(x=NMDS1,y=NMDS2,label=gsub(" ","\n", specieslabels)),
             size=6,alpha=0.5) +
   geom_point(data=dens_sites,aes(x=NMDS1,y=NMDS2,shape=mpa,color=mpa),size=3) +
   #geom_text(data=dens_sites,aes(x=NMDS1,y=NMDS2,label=site),size=6,vjust=0) +
   stat_ellipse(data=dens_sites, aes(x=NMDS1,y=NMDS2,color=mpa), 
                level = 0.95) +   
   stat_ellipse(data=dens_sites, aes(x=NMDS1,y=NMDS2,color=inout), 
                level = 0.95) +
   scale_color_manual("", values=allcolors,
                      breaks=c("MCD","VICRNM","IN","OUT"),
                      labels=c("MCD","VICRNM","IN","OUT")) +
   scale_shape_manual("", values = c(16,17)) +
     scale_x_continuous(limits=c(-1.25,1.5)) +
   theme(text=element_text(size=16), panel.background = element_blank(),
            axis.line=element_line())
```

We can follow up the NMDS with a PERMANOVA.

```{r groupDensTest}
dens_perm <- adonis2(density_matrix ~ paste(MPA,InOut), data = groupsitesum)
dens_perm
perm_p <- dens_perm$`Pr(>F)`[1]
```
Locations (in terms of MPA and in/out) are different in their density of groups of interest (df=`r dens_perm$Df[1]`, p=`r perm_p`). Since the PERMANOVA was significant, we can do a SIMPER analysis to see which groups are contributing the most to differences.

```{r groupDensTest2}
dens_sim <- with(groupsitesum, simper(density_matrix, paste(MPA,InOut)))
summary(dens_sim)
dens_sim
```

SIMPER analysis showed the largest contributor to these differences is small-bodied prey, accounting for 78-84% of the differences between locations.

### NMDS for biomass of groups

The code below creates a matrix of biomass of groups of interest at each site, using that to create an NMDS and the associated stressplot.

```{r groupBio}
biomass_matrix <- as.matrix(groupsitesum[paste0(names(grouplabels),".biomass")])
dimnames(biomass_matrix) <- list(as.character(groupsitesum$SiteID), 
                                 paste0(names(grouplabels),".biomass"))
bio_NMDS <- metaMDS(biomass_matrix, distance="jaccard", k=2)
stressplot(bio_NMDS)
```

The following code converts the NMDS into a data frame and plots it.

```{r groupBioPlot, fig.width = 7.5, fig.height = 7, fig.cap="NMDS plot of biomass of various functional groups of interest by MPA. Groups were adapted from Green et al (2012). Ellipses represent 95% groupings by MPA (inside and outside combined for each MPA)."}
bio_sites <- as.data.frame(scores(bio_NMDS))
bio_sites$site <- rownames(bio_sites)
bio_sites$mpa <- groupsitesum$MPA
bio_sites$inout <- groupsitesum$InOut

bio_species <- as.data.frame(scores(bio_NMDS, "species"))
bio_species$species <- rownames(bio_species)
bio_species$specieslabels <- grouplabels

ggplot() + 
      geom_label(data=bio_species,aes(x=NMDS1,y=NMDS2,label=gsub(" ","\n", specieslabels)),
                size=6,alpha=0.5) +
      geom_point(data=bio_sites,aes(x=NMDS1,y=NMDS2,shape=mpa,color=mpa),size=3) +
      #geom_text(data=bio_sites,aes(x=NMDS1,y=NMDS2,label=site),size=6,vjust=0) +
   scale_color_manual("", values=allcolors,
                      breaks=c("MCD","VICRNM","IN","OUT"),
                      labels=c("MCD","VICRNM","IN","OUT")) +
      scale_shape_manual("", values = c(16,17)) +
      stat_ellipse(data=bio_sites, aes(x=NMDS1,y=NMDS2,color=mpa), 
                   level = 0.95) +
   stat_ellipse(data=bio_sites, aes(x=NMDS1,y=NMDS2,color=inout), 
                level = 0.95) +
      #scale_x_continuous(limits=c(-1.7,1.1))+
      theme(text=element_text(size=16), panel.background = element_blank(),
            axis.line=element_line())
```

We can follow up the NMDS with a PERMANOVA.

```{r groupBioTest}
bio_perm <- adonis2(biomass_matrix ~ paste(MPA,InOut), data = groupsitesum)
bio_perm
```

Since the PERMANOVA is not significant, we don't need to do a SIMPER.

## Density and biomass of trophic groups by site

First we need to calculate the density (#/100m^2^) and biomass (g/100m^2^) of each trophic group at each site.

```{r trophsiteSum}
trophsitesum <- dl %>%
     group_by(SiteID, MPA, InOut, Troph) %>%
     summarize(TotalNumber=sum(TotalNumber),
               TotalBiomass=sum(TotalBiomass), .groups="drop")
trophdens <- trophsitesum %>% 
     pivot_wider(id_cols=c(SiteID, MPA,InOut), 
                 names_from = Troph, values_from = TotalNumber,
                 values_fill = 0)
trophbio <- trophsitesum %>% 
     pivot_wider(id_cols=c(SiteID, MPA,InOut), 
                 names_from = Troph, values_from = TotalBiomass,
                 values_fill = 0)
```

This table is the first 6 rows of the resulting tables:

```{r trophSiteSumTabs, echo = FALSE}
kable(head(trophdens), digits=0, align = "c",
      col.names = c("Site", "MPA","Location", "Herbivore","Invertivore","Piscivore","Planktivore"),
      caption = "Total density (#/100m^2^) of trophic groups across sites (first 6 rows).")
kable(head(trophbio), digits=0, align = "c",
      col.names = c("Site", "MPA","Location", "Herbivore","Invertivore","Piscivore","Planktivore"),
      caption = "Total biomass (g/100m^2^) of trophic groups across sites (first 6 rows).")
```

### NMDS for density of trophic groups

```{r trophDens}
trophdensmatrix <- as.matrix(trophdens[,levels(as.factor(dl$Troph))])
dimnames(trophdensmatrix) <- list(as.character(trophdens$SiteID), 
                                  as.character(levels(as.factor(dl$Troph))))

trophdens_NMDS <- metaMDS(trophdensmatrix, distance="jaccard", k=2)
stressplot(trophdens_NMDS)
```

```{r trophDensPlot, fig.width = 7.5, fig.height = 7, fig.cap="NMDS plot of density of trophic groups by MPA. Ellipses represent 95% groupings by MPA (inside and outside combined for each MPA)."}
trophdens_sites <- as.data.frame(scores(trophdens_NMDS))
trophdens_sites$site <- rownames(trophdens_sites)
trophdens_sites$mpa <- trophdens$MPA
trophdens_sites$inout <- trophdens$InOut

trophdens_species <- as.data.frame(scores(trophdens_NMDS, "species"))
trophdens_species$species <- rownames(trophdens_species)

ggplot() + 
      geom_label(data=trophdens_species,aes(x=NMDS1,y=NMDS2,label=species),
                size=6,alpha=0.5) +
      geom_point(data=trophdens_sites,aes(x=NMDS1,y=NMDS2,shape=mpa,color=mpa),size=3) +
   scale_color_manual("", values=allcolors,
                      breaks=c("MCD","VICRNM","IN","OUT"),
                      labels=c("MCD","VICRNM","IN","OUT")) +
      scale_shape_manual("", values = c(16,17)) +
      stat_ellipse(data=trophdens_sites, aes(x=NMDS1,y=NMDS2,color=mpa), 
                   level = 0.95) +
   stat_ellipse(data=trophdens_sites, aes(x=NMDS1,y=NMDS2,color=inout), 
                level = 0.95) +
      theme(text=element_text(size=16), panel.background = element_blank(),
            axis.line=element_line())

```

We can follow up the NMDS with a PERMANOVA.

```{r trophDensTest}
trophdens_perm <- adonis2(trophdensmatrix ~ paste(MPA,InOut), data = trophdens)
trophdens_perm
trophdensperm_p <- trophdens_perm$`Pr(>F)`[1]

```

Locations (in terms of MPA and in/out) are different in their density of trophic groups (df=`r trophdens_perm$Df[1]`, p=`r trophdensperm_p`). Since the PERMANOVA was significant, we can do a SIMPER analysis to see which groups are contributing the most to differences.
```{r trophDensTest2}
trophdenssim <- with(trophdens, simper(trophdensmatrix, paste(MPA,InOut)))
trophdenssim
```

Herbivores and invertivores contribute the most to differences in fish density across sites.

### NMDS for biomass of trophic groups

```{r trophBio}
trophbiomatrix <- as.matrix(trophbio[,levels(as.factor(dl$Troph))])
dimnames(trophbiomatrix) <- list(as.character(trophbio$SiteID), 
                                  as.character(levels(as.factor(dl$Troph))))

trophbio_NMDS <- metaMDS(trophbiomatrix, distance = "jaccard", k=2)
stressplot(trophbio_NMDS)

```

```{r trophBioPlot}
trophbio_sites <- as.data.frame(scores(trophbio_NMDS))
trophbio_sites$site <- rownames(trophbio_sites)
trophbio_sites$mpa <- trophbio$MPA
trophbio_sites$inout <- trophbio$InOut

trophbio_species <- as.data.frame(scores(trophbio_NMDS, "species"))
trophbio_species$species <- rownames(trophbio_species)

ggplot() + 
   geom_label(data=trophbio_species,aes(x=NMDS1,y=NMDS2,label=species),
             size=6,alpha=0.5) +
   geom_point(data=trophbio_sites,aes(x=NMDS1,y=NMDS2,shape=mpa,color=mpa),size=3) +
   #geom_text(data=trophbio_sites,aes(x=NMDS1,y=NMDS2,label=site),size=6,vjust=0) +
   scale_color_manual("", values=allcolors,
                      breaks=c("MCD","VICRNM","IN","OUT"),
                      labels=c("MCD","VICRNM","IN","OUT")) +
   scale_shape_manual("", values = c(16,17)) +
   stat_ellipse(data=trophbio_sites, aes(x=NMDS1,y=NMDS2,color=mpa), 
                level = 0.95) +
   stat_ellipse(data=trophbio_sites, aes(x=NMDS1,y=NMDS2,color=inout), 
                level = 0.95) +
   theme(text=element_text(size=20), panel.background = element_blank(),
         axis.line=element_line())
```

We can follow up the NMDS with a PERMANOVA.

```{r trophBioTest}
trophbio_perm <- adonis2(trophbiomatrix ~ paste(MPA,InOut), data = trophbio)
trophbio_perm
```

Locations (in terms of MPA and in/out) are different in their density of trophic groups (df=`r trophdens_perm$Df[1]`, p=`r trophdensperm_p`). Since the PERMANOVA was significant, we can do a SIMPER analysis to see which groups are contributing the most to differences.

```{r trophBioTest2}
trophbiosim <- with(trophbio, simper(trophbiomatrix, paste(MPA,InOut)))
trophbiosim
```

Invertivores, piscivores, and herbivores all contribute significantly to differences in biomass across sites.

## Commercially Important Species
```{r comSiteSum}
comsum <- dl %>%
     filter(Commercial=="Y") %>%
     group_by(SiteID, MPA, InOut, ScientificName) %>%
     summarize(TotalNumber=sum(TotalNumber),
               TotalBiomass=sum(TotalBiomass), .groups="drop")
comdens <- comsum %>%
     pivot_wider(id_cols=c(SiteID, MPA,InOut), 
                 names_from = ScientificName, values_from = TotalNumber,
                 values_fill = 0)
```
```{r comSiteTab, echo=FALSE}
kable(head(comdens), digits=0, align = "c",
      caption = "Mean density (#/100m^2^) of commercially-important species across sites (first 6 rows).")
```

There are too many zeros in this dataset for an NMDS to work.