---
title: "Lionfish"
author: "Sarah Heidmann"
date: "9/9/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

```
```{r library, include=FALSE}
# Load libraries
library(tidyverse)
library(knitr)
```
```{r dataImport, include=FALSE}
# Import data
lionsize <- read_csv("data/DeepLion_lionsizes.csv") %>% 
     mutate(Zm_pos = -Zm)
# Set colors
mpaColors <- c("#E69F00", "#56B4E9")
inoutColors <- c("#00BE6C","#F8766D")
```

## Summary

This document examines patterns in the density and biomass of the invasive lionfish, *Pterois volitans*. Here we look at presence and size distributions inside and outside the boundaries of the Hind Bank Marine Conservation District (MCD) and Virgin Islands Coral Reef National Monument (VICRNM), and density and biomass of lionfish as compared to that of native prey and predators.

## Size Frequency

```{r sizefreq, fig.width = 7.5, fig.height = 7, fig.cap="Size frequency distribution of lionfish, separated by MPA and inside/outside MPA boundaries."}
lionsize %>% 
   mutate(size=as.factor(size)) %>% 
   add_column(count=1) %>%
   complete(size, nesting(MPA, InOut), fill=list(count=0)) %>%
   group_by(MPA,InOut,size) %>% 
   summarize(total = sum(count), .groups="drop") %>%
ggplot(data=., aes(x=size, y=total,fill = InOut)) +
   geom_bar(color = "black",position = "dodge", stat="identity") +
   scale_fill_manual("",values = inoutColors) +
   xlab("Size (cm)") +
   ylab("Frequency") +
   scale_x_discrete(labels = c("6-10","11-20","21-30","31-40")) +
   scale_y_continuous(expand=c(0,0)) +
   facet_wrap(~MPA,ncol=1) +
   theme(panel.background = element_blank(), axis.line=element_line(),
         text = element_text(size = 16), legend.position = "bottom")
```

Lionfish were most abundant in the 20-30 cm size class.

## Depth Patterns

### Size with Depth
```{r depthSize, fig.width = 7.5, fig.height = 7, fig.cap="Total length of all observed lionfish by depth at which each was observed, separated by MPA and inside/outside MPA boundaries."}
sizedepthlm <- lm(size~Zm_pos, lionsize)

ggplot(data = lionsize, aes(x=Zm_pos,y=size)) +
      geom_point(size = 4, aes(shape=MPA, color=InOut)) +
   geom_smooth(formula = y~x, method = "lm", group=1) + # for MPA
   scale_color_manual("",values = inoutColors) +
   scale_shape_manual("MPA", values = c(16,17)) +
      xlab("Depth (m)") +
      ylab("Lionfish Size (cm)") +
      scale_x_continuous(breaks = seq(10,50,10), labels = seq(10,50,10),
                         limits = c(15,50))+
      scale_y_continuous(breaks = seq(0,50,10), labels = seq(0,50,10),
                         limits = c(0,50),expand = c(0,0)) +
      theme(panel.background= element_blank(), axis.line=element_line(),
            legend.key = element_blank(), text = element_text(size = 16))

summary(sizedepthlm)
```
There was a suggestive relationship between depth and lionfish size (p=0.09), with larger lionfish tending to occur deeper. A more accurate relationship may be able to be determined with a higher-resolution size classification of lionfish.

### Density with Depth

```{r depthDensity, fig.width = 7.5, fig.height = 7, fig.cap="Number of lionfish seen at each site by depth of the site, separated by MPA and inside/outside MPA boundaries.", message=FALSE}
numdepthsum <- lionsize %>% 
     group_by(MPA, InOut, Zf, Zm_pos) %>%
     summarize(density = length(size), .groups = "drop")

ggplot(data=numdepthsum, aes(x=Zm_pos,y=density)) +
      geom_point(size = 4, aes(shape = MPA, color = InOut)) +
      #geom_smooth(formula=y~x,method = "lm", group=1) +
   scale_color_manual("",values = inoutColors) +
   scale_shape_manual("MPA", values = c(16,17)) +
   xlab("Depth (m)") +
   ylab(expression(paste("Lionfish Density (#/100",m^2,")"))) +
   scale_x_continuous(breaks = seq(20,50,10), labels = seq(20,50,10))+
   scale_y_continuous(limits = c(0,8), breaks = seq(0,8,2), labels = seq(0,8,2),
                      expand = c(0,0))+
      theme(panel.background= element_blank(), axis.line=element_line(), 
            text = element_text(size = 16))
```

There was no apparent relationship between lionfish density and depth.

### Biomass with Depth
```{r depthBiomass, fig.width = 7.5, fig.height = 7, fig.cap="Biomass of lionfish seen at each site by depth of the site, separated by MPA and inside/outside MPA boundaries.", message = FALSE}
biodepthsum <- lionsize %>%
     group_by(MPA, InOut, Zf,Zm_pos) %>%
     summarize(biokg = sum(bio)/1000, .groups="drop")

ggplot(data=biodepthsum, aes(x=Zm_pos,y=biokg,shape=MPA,color=InOut)) +
      geom_point(size =4, aes(color=InOut)) +
      #geom_smooth(method= "lm", group=1) +
   geom_smooth(formula = y~x, method="lm",se=FALSE)+
   scale_color_manual("",values = inoutColors) +
   scale_shape_manual("MPA", values = c(16,17)) +
   xlab("Depth (m)") +
   ylab(expression(atop("Lionfish Biomass", 
                        paste("(kg/100",m^2,")")))) +
   scale_x_continuous(breaks = seq(20,50,10), labels = seq(20,50,10))+
   scale_y_continuous(limits = c(0,3), breaks = seq(0,3,0.5),
                      labels = seq(0,3,0.5), expand = c(0,0))+
      theme(panel.background= element_blank(), axis.line=element_line(), 
            text = element_text(size = 16))
```
```{r depthBiomassTest, echo = TRUE}
summary(lm(biokg ~ Zm_pos,
           filter(biodepthsum, MPA=="MCD" & InOut=="IN")))
```

There is a significant positive relationship between depth and lionfish biomass within the MCD (p=0.03), which is where many of the lionfish were found (`r round(nrow(filter(lionsize, MPA=="MCD" & InOut=="IN"))/nrow(lionsize)*100,0)`%). There was no relationship, however, between depth and lionfish density, suggesting that larger lionfish tend to exist on deeper reefs.